---
title: 'STAT 510 HW #3'
author: "Steve Harms"
date: "January 23, 2018"
output:
  word_document: default
  pdf_document: default
  html_document:
    fig_height: 4
    fig_width: 4
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include = FALSE}
library(Sleuth3)
library(ggplot2)
library(MASS)
library(tidyverse)
```
#1)
##a)
```{r}
df <- case0501
g <- ggplot(data = df, aes(x = Diet, y = Lifetime))
g + geom_boxplot()
```
##b)
```{r}
mod <- lm(Lifetime~0+Diet, data = df)
anva <- anova(mod)
anva
anva$`Sum Sq`[2]

#alternative using matrices
lenx <- length(df$Diet)
diets <- arrange(df, Diet)
Ident <- diag(lenx)
X <- model.matrix(~0+factor(diets$Diet))
Px <- X%*%ginv(t(X)%*%X)%*%t(X)
y <- diets$Lifetime
sse <- t(y)%*%(Ident-Px)%*%y
sse
```
##c)
```{r}
#Using ANOVA Table
anva
anva$`Mean Sq`[2]

#using matrices
sig2hat <- sse/(lenx - 6)
sig2hat
```
##d)
```{r}
#using lm
dietsn <- diets
#recode
dietsn$Diet[dietsn$Diet=="N/R50"] <-  "lopro"
#lm
mod.red <- lm(Lifetime~0+Diet, data = dietsn)
anva.red <- anova(mod.red)
anva.red
anva.red$`Sum Sq`[2]

#using matrices
Xo <- model.matrix(~0+factor(dietsn$Diet))
Po <- Xo%*%ginv(t(Xo)%*%Xo)%*%t(Xo)
y <- diets$Lifetime
sse.red <- t(y)%*%(Ident-Po)%*%y
sse.red
```
##e)
```{r}
#using anova()
anova(mod.red, mod)$F[2]
#using matrices
Fstat <- ((sse.red - sse)/(anva.red$Df[2]-anva$Df[2])) / (sse/anva$Df[2])
Fstat
#and a p-value as well
pf(Fstat, df1 = (anva.red$Df[2]-anva$Df[2]), df2 = (anva$Df[2]), lower.tail = F)
```
##f)
The F-statistic can be used to test whether there is a _significant difference in mean lifetimes_ of the two groups of mice that were fed reduced calorie diets only after weaning but not before (groups N/R50 and N/R50 lopro). The test would help determine whether there was an effect of the low protein diet when mice were fed reduced calorie diets only after weaning.
##g)
Our $\beta$ matrix is just a vector of group means, 
$$\mathbf{\beta} = \left[\begin{array}{rrr}
\mu_1 \\ \mu_2 \\ \mu_3 \\ \mu_4 \\ \mu_5 \\ \mu_6 \\
\end{array}\right]
$$
Where each $\mu_i$ is the group mean for each treatment group listed in the order of the statement. We want to test whether group 3 has the same treatment mean as group 5, or that $\mu_3 - \mu_5 = 0$. Then
$$\mathbf{C} = \left[\begin{array}
{rrr}
0 &0 & 1 & 0 & -1 & 0 \\
\end{array}\right]
$$
and $\textbf{d} = 0$

Then our F-statistic should be the same as before, and now using the matrix method on slide 20, we get
```{r}
C = t(c(0,0,1,0,0,-1)) #I sorted the data so C is different than specified above
q = 1
d = 0
y <- diets$Lifetime
X <- model.matrix(~0+factor(diets$Diet))
betahat = ginv(t(X)%*%X)%*%t(X)%*%y
Fc = (t((C%*%betahat))%*%solve(C%*%(ginv(t(X)%*%X))%*%t(C))%*%(C%*%betahat) / 1 ) / (sig2hat)
Fc
```

#2)
Let $$\mathbf{A} = \left[\begin{array}
{rrr}
-1 & 1 \\
 1 & 1\\
\end{array}\right]$$
and let $$\mathbf{G} = \left[\begin{array}
{rrr}
-1/2 & -1 \\
 1 & -1/2\\
\end{array}\right]$$
Simple matrix multiplication will show that $AGA = A$, thus by definition G is a generalized inverse of $A$. But $A$ is symmetric (off-diagonal elements are the same), and $G$ is clearly not symmetric as $-1 \neq 1$, so generalized inverses of symmetric matrices need not be symmetric.

End.










