---
title: "Homework 5"
author: "Steve Harms"
date: "April 22, 2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include = FALSE}
library(spatstat)
library(RColorBrewer)
library(ggplot2)
library(viridisLite)
```
#2)
##a)
```{r, include = FALSE}
bramble <- scan(file="brambles3.txt", sep = ',')
bramble <- bramble[!is.na(bramble)]
length(bramble) ; tail(bramble)
l <- seq(1:718); eventries <- which(l%%2==0)
x <- 9*bramble[-eventries]
y <- 9*bramble[eventries]
brambles <- as.data.frame(cbind(x,y))
```
```{r}
canes <- ppp(x, y, window=owin(c(0,9),c(0,9)))

#a plot of the data
plot(x = x, y = y)

#plot of the density using Diggle's Bandwidth
image(density(canes, diggle=T), main = "Estimated Intensity", col = brewer.pal(n = 9, name = "Oranges"))
```

**It appears there may be some clustering based on the plot of the intensity, but the plot of the original data is inconclusive.**

##b)
```{r}
#K-function
Kc <- Kest(canes, correction="Ripley")
plot(Kc, main="Estimated K function")
plot(Kc,  iso - theo ~ r, main="Deviation from CSR (K-function)")
abline(h=0, lty=2, col="gray40")
#L-function
Lc <- Lest(canes)
plot(Lc, main = "Estimated L Function")
plot(Lc,  iso - theo ~ r, main="Deviation from CSR (L-Function)")
abline(h=0, lty=2, col="gray40")
#values are large near 0, indication of some clustering
```

**Both the K- and L- functions have positive values near 0, indicating that the plot is clustered. However, note that the deviation  from CSR for the L- function is actually less than 0 for very small distance from origin, which indicates that this process may not actually be clustered.**

#3)
##a)
```{r, include=FALSE}
redwoods <- scan(file = "redwood.txt")
l.r <- seq(1:length(redwoods)); eventries.r <- which(l.r%%2==0)
x.r <- 23*redwoods[-eventries.r] ; y.r <- 23*redwoods[eventries.r];
redwood <- as.data.frame(cbind(x.r, y.r))
```
```{r}
trees <- ppp(x.r, y.r, window=owin(c(0,23),c(0,23)))
#a plot of the data
plot(x = x.r, y = y.r)
#plot of the density
plot(density(trees, diggle=T), main = "Estimated Intensity", col = plasma(10))
```

##b)
```{r}
#K-function
Ks <- Kest(trees, correction="Ripley")
plot(Ks, main="Estimated K function")
plot(Ks,  iso - theo ~ r, main="Deviation from CSR (K)")
abline(h=0, lty=2, col="gray40")
#L-function
Ls <- Lest(trees)
plot(Ls, main = "Estimated L Function")
plot(Ls,  iso - theo ~ r, main="Deviation from CSR (L)")
abline(h=0, lty=2, col="gray40")
```

**While our intuition suggests that this process should be clustered around trees, the K- and L-functions and actually negative near 0, indicating that this process is actually somewhat regular/unclustered.**
